---
title: "Bagging and boosting of RVFL neural networks"
author: "Søren B. Vilsen"
date: "5. March 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bagging and boosting of RVFL neural networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library("BRVFL")
```

## Introduction
The `BRVFL`-package is an `R` implementation of Bootstrap aggregated (Bagged) and boosted Random Vector Functional Link (RVFL) Neural Network.

## Random Vector Functional Link Neural Networks
The idea of RVFL is to simplify a feed forward neural network, by only estimating the weights connecting the last hidden layer to the output layer, while the remaining weights are sampled randomly from a bounded distribution. The idea was introduced in [], and then later rediscovered, and re-branded to extreme learning machines (ELMs), in []. The only difference between ELM and RVFL, is that RVFL also allows for connections between the input and output layers making it a more general framework. The general structure of an RVFL network can be seen in Fig. [].

The 

## Bagging RVFL Neural Networks
The general idea of bagging is to stabalise predicted response by re-sampling from the data (with replacement), estimate the parameters of the model (in this case an RVFL), creating a series of models. Given a new data point ensemble predictions can be made using the estimated models.  

## Boosting RVFL Neural Networks


## References